---
title: "Homework 2"
author: "Piotr Mankowski"
date: "January 19, 2018"
output:
  html_notebook: default
  pdf_document: default
---


```{r setup}
library(data.table)
library(uwIntroStats)
library(regress)

mri = as.data.table(read.table("../data/mri.txt", header=TRUE, stringsAsFactors = FALSE))
mri[, obstime_yr:=obstime/365]
mri[, dead_at_5:=ifelse((obstime_yr <= 5.0 & death == 1), 1, 0), by='ptid']

```

## Question 1
*Perform a statistical analysis evaluating an association between **serum creatinine level** and **5 year all-cause mortality** by comparing **geometric mean serum creatinine levels** between groups defined by vital status at 5 years.   In your analysis, allow for heteroscedasticity.  Provide full statistical inference about an association between creatinine and 5 year all-cause mortality.*

Basically, we're examining the association between `crt` and `dead_at_5yr` using geometric means (like in the discussion). 

```{r}

```

### Methods
### Inference

#### a) What are the sample size, sample mean and sample standard deviation of serum creatinine levels among subjects who survived at least 5 years? 
